The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-07 Fsing/FThresholdsPaper/FThresholds.tex
Minor edits to the paper

This paper describes the \emph{Macaulay2} \cite{M2} package \emph{FrobeniusThresholds} \cite{FThresholdsPackage}, which provides tools for computing certain fundamental invariants arising in positive characteristic commutative algebra.  More precisely, this package concerns the effective computation of \emph{$F$-pure thresholds}, \emph{$F$-thresholds}, and \emph{$F$-jumping exponents}.

@@

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper,M2}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived (and have greatly grown) from an older package \emph{PosChar}.

@@

The overarching goal of this paper is paper is to describe and illustrate the main functionality implemented in the \emph{FrobeniusThresholds} package.  We break this down into sections, and then conclude by pointing out some potential future directions for the package.

@@

\subsection*{Acknowledgements.}  First and foremost, we enthusiastically thank everyone who helped complete the \emph{FrobeniusThresholds} package. 
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.

@@

Finally, we are grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its generous support for our 2019 Coding Sprint.
The current version of the package was finalized during these events.

@@

We first consider the {\tt nu} function, a fundamental component of the package \emph{FrobeniusThresholds}.   As in the introduction, in this section we work in a polynomial ring over a finite field of characteristic $p>0$.  As in the introduction, $\idealm$ will denote the ideal of this ring generated by the variables.

Fix a natural number $e$.  If $f$ is a nonzero polynomial contained in the radical of an ideal $J$, then the command {\tt nu(e,f,J)} outputs the integer $\nu_f^J(p^e)$ defined in the introduction, and
%, the maximal integer exponent such that $f^n$ is not contained in $J^{[p^e]}$.
more generally, if a nonzero ideal $I$ is contained in the radical of $J$, then {\tt nu(e,I,J)} outputs $\nu_I^J(p^e)$. Consider the following example.  %, the maximal integer exponent $n$ such that $I^n$ is not contained in $J^{[p^e]}$.

@@

As described in the introduction, the integer $\nu_I^J(p^e)$ is defined in terms of containment conditions on powers of ideals.
For positive integers $n$ and $e$, and an ideal $I$ contained in the radical of another ideal $J$, we have that
\begin{equation*}
  I^n \subseteq J^{[p^e]} \Longleftrightarrow (I^n)^{[1/p^e]} \subseteq J
\end{equation*}
where $(I^n)^{[1/p^e]}$ denotes the $p^e$-th \emph{Frobenius root} of $I^n$, as defined in \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds}.
The option {\tt ContainmentTest} for {\tt nu} allows the user to choose which of the two types of containment statements appearing above to use toward the calculation of $\nu_I^J(p^e)$.

@@

The option {\tt ContainmentTest} has a third possible value, {\tt FrobeniusPower}, which may produce a different result than the other two values.
Given an ideal $I$, the first, third, and fourth authors introduced the notion of a (generalized) Frobenius power $I^{[n]}$ of $I$, when $n$ is an arbitrary positive integer, not necessarily of the form $n = p^e$ \cite{hernandez+etal.frobenius_powers}.
When {\tt ContainmentTest} is set to {\tt FrobeniusPower}, rather than  computing $\nu_I^J(p^e)$, the function {\tt nu} computes the maximal integer $n$ for which $I^{[n]}$ is not contained in $J^{[p^e]}$.  This number is denoted $\mu_I^J(p^e)$ in \emph{loc.\ cit.}, and equals $\nu_I^J(p^e)$ when $I$ is replaced with a polynomial, or is a principal ideal, but need not agree in general.
For example, the two commands below give different output values.

@@

\noindent In line with the philosophy described in the introduction, after appropriately normalizing and taking the limit as $e$ approaches infinity, the values $\mu_I^J(p^e)$ determine certain invariants of interest called \emph{critical exponents}.

The last option we describe for {\tt nu} is {\tt IsLocal}.
Notice that $\nu_I^\idealm(p^e)$ can be interpreted as the maximum integer $n$ for which $(I^n)^{[1/p^e]}$ is not contained in $\idealm$. 
When computing its value globally, it is the maximum integer $n$ for which $(I^n)^{[1/p^e]} \neq R$, and when 
the value of {\tt IsLocal} is set to {\tt false} (from its default value {\tt true}), then {\tt nu} instead determines this value $n$.  Note that this value is also the minimum $\nu_I^\mathfrak{n}(p^e)$ among all maximal ideals $\mathfrak{n}$ of $R$.

@@

The above noted algorithm for computing the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and {\tt fpt} can sometimes hang when attempting that factorization, when the factorization occurs in a Galois field of excessively large order.
This is a situation when the user will want to set the option \texttt{UseSpecialAlgorithms} to \texttt{false}.
However, when a factorization is already known, instead of passing {\tt f} to {\tt fpt}, the user can pass a list containing all the pairwise coprime linear factors of $f$ to {\tt fpt}, and a list containing their respective multiplicities.

>>>>

This paper describes the \emph{Macaulay2} package \emph{FrobeniusThresholds} \cite{M2, FThresholdsPackage}.  This package provides tools for computing certain fundamental invariants in positive characteristic commutative algebra.  More precisely, this package concerns the effective computation of \emph{$F$-pure thresholds}, \emph{$F$-thresholds}, and \emph{$F$-jumping exponents}.

@@

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived from, and have greatly expanded, the older package \emph{PosChar}.

@@

The goal of this paper is paper is to describe and illustrate the main functionality implemented in the \emph{FrobeniusThresholds} package.  We describe the organization of this paper below.

@@

\subsection*{Acknowledgements.}  The authors enthusiastically thank everyone who helped complete the \emph{FrobeniusThresholds} package. 
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.

@@

Finally, the authors are grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its generous support for our 2019 Coding Sprint.
The current version of the package was finalized during these events.

@@

We first consider the {\tt nu} function, a fundamental component of the package \emph{FrobeniusThresholds}.   In this section, we adopt the notation established in the introduction.  Namely, we work in a polynomial ring over a finite field of characteristic $p>0$,  $\idealm$ denotes the ideal generated by the variables, and $e$ is a natural number.

If $f$ is a nonzero polynomial contained in the radical of an ideal $J$, then the command {\tt nu(e,f,J)} outputs the integer $\nu_f^J(p^e)$ defined in the introduction, and
%, the maximal integer exponent such that $f^n$ is not contained in $J^{[p^e]}$.
more generally, if a nonzero ideal $I$ is contained in the radical of $J$, then {\tt nu(e,I,J)} outputs $\nu_I^J(p^e)$. Consider the following example.  %, the maximal integer exponent $n$ such that $I^n$ is not contained in $J^{[p^e]}$.

@@

As described in the introduction, the integer $\nu_I^J(p^e)$ is the maximal integer $n$ such that the $n$-th power of $I$ does not lie in the $p^e$-th Frobenius power of $J$.  However, 
\begin{equation*}
  I^n \subseteq J^{[p^e]} \Longleftrightarrow (I^n)^{[1/p^e]} \subseteq J
\end{equation*}
where $(I^n)^{[1/p^e]}$ denotes the $p^e$-th \emph{Frobenius root} of $I^n$, as defined in \cite{BlickleMustataSmithDiscretenessAndRationalityOfFThresholds}.
The option {\tt ContainmentTest} for {\tt nu} allows the user to choose which of the two types of containment statements appearing above to use toward the calculation of $\nu_I^J(p^e)$.

@@

The option {\tt ContainmentTest} has a third possible value, {\tt FrobeniusPower}, which may produce a different result than the other two values.
Given an ideal $I$, the first, third, and fourth authors introduced the notion of a (generalized) Frobenius power $I^{[n]}$ of $I$, when $n$ is an arbitrary positive integer, not necessarily of the form $n = p^e$ \cite{hernandez+etal.frobenius_powers}.
When {\tt ContainmentTest} is set to {\tt FrobeniusPower}, rather than  computing $\nu_I^J(p^e)$, the function {\tt nu} computes the maximal integer $n$ for which $I^{[n]}$ is not contained in $J^{[p^e]}$.  This number is denoted $\mu_I^J(p^e)$ in \emph{loc.\ cit.}, and equals $\nu_I^J(p^e)$ when $I$ is replaced with a polynomial, or is a principal ideal.  However, these number need not agree in general, as we see below.

@@

 In line with the philosophy described in the introduction, after appropriately normalizing and taking the limit as $e$ approaches infinity, the values $\mu_I^J(p^e)$ determine certain invariants of interest called \emph{critical exponents}.

The last option we describe for {\tt nu} is {\tt IsLocal}.
Recall that $\nu_I^\idealm(p^e)$ can be interpreted as the maximum integer $n$ for which $(I^n)^{[1/p^e]}$ is not contained in $\idealm$. 
When computing its value globally, it is the maximum integer $n$ for which $(I^n)^{[1/p^e]} \neq R$.  When 
the value of {\tt IsLocal} is set to {\tt false} (from its default value {\tt true}), then {\tt nu} instead determines this maximal integer $n$, which can also be characterized as the minimal integer $\nu_I^\mathfrak{n}(p^e)$ as $\mathfrak{n}$ varies among all maximal ideals of $R$.

@@

The above noted algorithm for computing the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and {\tt fpt} can sometimes hang when attempting that factorization, when the factorization occurs in a Galois field of excessively large order.
This is a situation when the user will want to set the option \texttt{UseSpecialAlgorithms} to \texttt{false}.
However, when a factorization is already known, instead of passing $f$ to {\tt fpt}, the user can pass a list containing all the pairwise coprime linear factors of $f$ to {\tt fpt}, and a list containing their respective multiplicities.
